# Predicting and visualizing breast cancer diagnosis based on available numerical data

## Data description
Each row of the data corresponds to breast cancer cells of a patient. Each column contains measurement data of all the cells of a patient and one column contains the diagnosis (Malignant/Benign) of the patient.

The database contains all numerical variables except one categorical variable: diagnosis
```{r}
'Please uncomment and install packages as neccessary'

'install.packages("randomForest")
install.packages("Amelia")
install.packages("caTools")
install.packages("rpart.plot")
install.packages("gridExtra")
install.packages("dplyr")
install.packages("stringr")
install.packages("rmarkdown")
install.packages("caTools")
install.packages("rpart.plot")
install.packages("corrplot")
install.packages("Hmisc")
install.packages("class")
install.packages("e1071")
install.packages("neuralnet")
install.packages("arm")
install.packages("ggpmisc")'

library(gridExtra)
library(randomForest)
library(Amelia)
library(dplyr)
library(ggplot2)
library(rmarkdown)
library(caTools)
library(stringr)
library(readr) 
library(rpart.plot)
library(corrplot)
library(Hmisc)
library(class)
library(e1071)
library(neuralnet)
library(arm)
library(ggpmisc)

```

## Displaying Data
```{r}
'reading the data file'
cancer.df <- read.csv("data.csv", header = TRUE)

'displaying the dataset'
head(cancer.df)
describe(cancer.df)
missmap(cancer.df, main='Raw data', col=c("red","blue"), legend = FALSE)

```

## Cleaning Data
```{r}
cancer.df <- cancer.df[-c(33)] 
head (cancer.df)
any(is.na(cancer.df))
missmap(cancer.df, main='Cleaned data', col=c("red","blue"), legend = FALSE)
cancer.df$diagnosis <- as.factor(cancer.df$diagnosis)

```

## Analysing and Visualizing data
```{r}
'Corelation matrix HeatMap'
corr_mat <- cor(cancer.df[,3:12])
corrplot::corrplot(corr_mat, order = "hclust", tl.cex = 1, addrect = 8)

```

### Sample graphs showing details of Malignant and Benign cells having high correlation
```{r}
ggplot(cancer.df,
       aes(area_mean, perimeter_mean)) + 
       geom_point(aes(color=factor(diagnosis)),alpha=0.3) + 
       scale_fill_discrete(name="diagnosis", breaks=c("0", "1"),
                           labels=c("Malignant", "Benign")) + 
       labs(title = "Area_Mean vs Perimeter_Mean")

ggplot(cancer.df,
       aes(smoothness_se, symmetry_mean)) +
       geom_point(aes(color=factor(diagnosis)),alpha=0.3) +
       scale_fill_discrete(name="diagnosis", breaks=c("0", "1"),
                           labels=c("Malignant", "Benign")) + 
       labs(title = "Smoothness_Standard_Error vs Symmetry_Mean")

```

### Distribution of Radius_Mean of Malignant annd benign cells
```{r}
'splitting malignant and benign data'
benign.df <- cancer.df[cancer.df$diagnosis == "B",]
malign.df <- cancer.df[cancer.df$diagnosis == "M",]

ggplot(cancer.df, 
       aes(x=radius_mean, color= diagnosis, fill = diagnosis, alpha = 0.3)) +                                      geom_histogram(binwidth = 0.1) +
       geom_vline(data=malign.df, aes(xintercept=mean(radius_mean), 
                                      color = diagnosis), 
                                      linetype="dashed")+
       geom_vline(data=benign.df, aes(xintercept=mean(radius_mean),
                                      color = diagnosis), 
                                      linetype="dashed")

```


```{r}
'comparing means and outliers with boxplot'
ggplot(cancer.df,
       aes(x=diagnosis, y=radius_mean, fill = diagnosis)) +
      geom_boxplot() +
      xlab("Diagnosis")+
      ylab("Radius Mean")
  
```
### Summary of distributions of significant variables based on correlation
```{r}

p1 <- ggplot(cancer.df, aes(x=diagnosis, y=radius_mean, 
                      fill = diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Radius Mean")+
             ggtitle("Distribution of Radius mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(radius_mean)), 
                                      linetype="dashed")

p2 <- ggplot(cancer.df, aes(x=diagnosis, y=texture_mean, 
                      fill = diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Texture Mean")+
             ggtitle("Distribution of Texture Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$texture_mean)), 
                                      linetype="dashed")

p3 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$perimeter_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Perimeter Mean")+
             ggtitle("Distribution of Perimeter Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$perimeter_mean)), 
                                      linetype="dashed")

p4 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$area_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Area Mean")+
             ggtitle("Distribution of Area Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$area_mean)), 
                                      linetype="dashed")

p5 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$smoothness_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Smoothness Mean")+
             ggtitle("Distribution of Smoothness Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$smoothness_mean)), 
                                      linetype="dashed")

p6 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$compactness_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Compactness Mean")+
             ggtitle("Distribution of Compactness Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$compactness_mean)), 
                                      linetype="dashed")

p7 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$concavity_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Concavity Mean")+
             ggtitle("Distribution of Concavity Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$concavity_mean)), 
                                      linetype="dashed")

p8 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$concave.points_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Concave Points Mean")+
             ggtitle("Distribution of Concave Points Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$concave.points_mean)), 
                                      linetype="dashed")

p9 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$symmetry_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Symmetry Mean")+
             ggtitle("Distribution of Symmetry Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$symmetry_mean)), 
                                      linetype="dashed")

p10 <- ggplot(cancer.df, aes(x=cancer.df$diagnosis, y=cancer.df$fractal_dimension_mean, 
                      fill = cancer.df$diagnosis)) +
             geom_violin() +
             geom_boxplot(width=0.1)+
             xlab("Diagnosis")+
             ylab("Fractal Dimension Mean")+
             ggtitle("Distribution of Fractal Dimension Mean for different diagnosis")+
             geom_hline(data=cancer.df, aes(yintercept=mean(cancer.df$fractal_dimension_mean)), 
                                      linetype="dashed")

p1
p2
p3
p4
p5
p6
p7
p8
p9
p10

```


```{r}
cancer.df <- read.csv("data.csv", header = TRUE)

cancer.df$X <- NULL
cancer.df$ids <- cancer.df$id
cancer.df$id <- NULL
cancer.df$diagnosis <- as.factor(cancer.df$diagnosis)
str(cancer.df)
```

## Understanding unacceptable error data
```{r}

'radius with standard error more than 10% of mean'
error.df <- cancer.df[cancer.df$radius_se > 0.1*mean(cancer.df$radius_mean),]
error.df$id <- 1:nrow(error.df)

ggplot(error.df, aes(x=id, y=radius_mean, fill=diagnosis)) +
  geom_bar(stat ="identity") +
  geom_errorbar( aes(x=id, y=radius_mean, ymin=radius_mean-radius_se,
                     ymax=radius_mean+radius_se), width=0.5, colour="pink",
                    alpha=0.9, size=1.5) +
  ggtitle("Radius standard error greater than 10% of mean") +
  theme(plot.title = element_text(hjust = 0.5)) 
  

'texture with standard error more than 10% of mean'
error.df <- cancer.df[cancer.df$texture_se > 0.1*mean(cancer.df$texture_mean),]
error.df$id <- 1:nrow(error.df)

ggplot(error.df, aes(x=id, y=texture_mean, fill=diagnosis)) +
  geom_bar(stat ="identity") +
  geom_errorbar( aes(x=id, y=texture_mean, ymin=texture_mean-texture_se,
                     ymax=texture_mean+texture_se), width=0.5, colour="pink",
                    alpha=0.9, size=1.5) +
  ggtitle("Texture standard error greater than 10% of mean") +
  theme(plot.title = element_text(hjust = 0.5))  


'perimeter with standard error more than 10% of mean'
error.df <- cancer.df[cancer.df$perimeter_se > 0.1*mean(cancer.df$perimeter_mean),]
error.df$id <- 1:nrow(error.df)

ggplot(error.df, aes(x=id, y=perimeter_mean, fill=diagnosis)) +
  geom_bar(stat ="identity") +
  geom_errorbar( aes(x=id, y=perimeter_mean, ymin=perimeter_mean-perimeter_se,
                     ymax=perimeter_mean+perimeter_se), width=0.5, colour="pink",
                    alpha=0.9, size=1.5) +
  ggtitle("Perimeter standard error greater than 10% of mean") +
  theme(plot.title = element_text(hjust = 0.5))

'area with standard error more than 10% of mean'
error.df <- cancer.df[cancer.df$area_se > 0.1*mean(cancer.df$area_mean),]
error.df$id <- 1:nrow(error.df)

ggplot(error.df, aes(x=id, y=area_mean, fill=diagnosis)) +
  geom_bar(stat ="identity") +
  geom_errorbar( aes(x=id, y=area_mean, ymin=area_mean-area_se,
                     ymax=area_mean+area_se), width=0.1, colour="pink",
                    alpha=0.9, size=0.8) +
  ggtitle("Area standard error greater than 10% of mean") +
  theme(plot.title = element_text(hjust = 0.5))

'smoothness with standard error more than 10% of mean'
error.df <- cancer.df[cancer.df$smoothness_se > 0.1*mean(cancer.df$smoothness_mean),]
error.df$id <- 1:nrow(error.df)

ggplot(error.df, aes(x=id, y=smoothness_mean, fill=diagnosis)) +
  geom_bar(stat ="identity") +
  geom_errorbar( aes(x=id, y=smoothness_mean, ymin=smoothness_mean-smoothness_se,
                     ymax=smoothness_mean+smoothness_se), width=0.5, colour="pink",
                    alpha=0.9, size=1) +
  ggtitle("Smoothness standard error greater than 10% of mean") +
  theme(plot.title = element_text(hjust = 0.5))

```

For variables (compactness, conacavity, symmetry...), the error margin can't be considered as a factor. Because these variables are relatively less adverse and almost all the data entails high standard error.

#Logistic Regression
```{r}
split <- sample.split(cancer.df$diagnosis, Split = 0.7)
training_set <- subset(cancer.df, split == T)
testing_set <- subset(cancer.df, split == F)

regression.df <- bayesglm(formula=diagnosis ~ . , family = binomial(link='logit'),
                     maxit = 100, data = training_set)

summary(regression.df)

probability.df <- predict(regression.df,newdata=testing_set[2:32],type='response')

fd <-table(testing_set$diagnosis, probability.df > 0.5)
fd

fd.t <-sum(diag(fd))/sum(fd)
print(fd.t)
```

#Random Forest
```{r}
rforest.df <- randomForest(diagnosis ~., data = training_set)

p.value <- predict(rforest.df, testing_set[2:32])
rd <- table(p.value, testing_set$diagnosis)
print(rd)

rd.t <-sum(diag(rd))/sum(rd)
print(rd.t)
```

#KNN
```{r}
normalize <- function(x){
  return((x-min(x))/(max(x)-min(x)))
}

cancer.df.k<- as.data.frame(lapply(cancer.df[2:30],normalize))
cancer.df.k$diagnosis <- cancer.df$diagnosis

split <- sample.split(cancer.df.k$diagnosis, Split = 0.7)
training_set_k <- subset(cancer.df.k, split == T)
testing_set_k <- subset(cancer.df.k, split == F)

p.cancer.df <- knn(training_set_k[1:29],testing_set_k[1:29],training_set_k$diagnosis,k=1)

mean(testing_set_k$diagnosis != p.cancer.df)

p.cancer.df <-NULL
er <-NULL

for(i in 1:10){
  p.cancer.df <- p.df<- knn(training_set_k[1:29],testing_set_k[1:29],training_set_k$diagnosis,k=i)
  er[i] <- mean(testing_set_k$diagnosis != p.cancer.df)
}

k.val <- 1:10
ed <- data.frame(er,k.val)

pl <- ggplot(ed,aes(x=k.val,y=er)) + geom_point()
pl + geom_line(lty=1,color='blue')

p.cancer.df <- knn(training_set_k[1:29],testing_set_k[1:29],training_set_k$diagnosis,k=5)
mean(testing_set_k$diagnosis != p.cancer.df)

c_mat <-table(testing_set_k$diagnosis, p.cancer.df)
print(c_mat)

c_mat.t <-sum(diag(c_mat))/sum(c_mat)
print(c_mat.t)
```

#Neural Network
```{r}
normalize <- function(x){
  return((x-min(x))/(max(x)-min(x)))
}

cancer.df.k<- as.data.frame(lapply(cancer.df[2:31],normalize))
cancer.df.k$diagnosis <- cancer.df$diagnosis
cancer.df.k$diagnosis <- as.numeric(cancer.df.k$diagnosis)

loop <- function(can){
  for(i in 1:length(can)){
    if(can[i] == 1){
      can[i] <- 0
    }else{
      can <- 1
    }
  }
  return(can)
}

cancer.df.k$diagnosis <-sapply(cancer.df.k$diagnosis,loop)

split <- sample.split(cancer.df.k$diagnosis, Split = 0.7)
training_set_k <- subset(cancer.df.k, split == T)
testing_set_k <- subset(cancer.df.k, split == F)

nn <- neuralnet(diagnosis ~ radius_mean + texture_mean + perimeter_mean + area_mean + 
                  smoothness_mean + compactness_mean + concavity_mean + concave.points_mean + 
                  symmetry_mean + fractal_dimension_mean + radius_se + texture_se + 
                  perimeter_se + area_se + smoothness_se + compactness_se + 
                  concavity_se + concave.points_se + symmetry_se + fractal_dimension_se + 
                  radius_worst + texture_worst + perimeter_worst + area_worst + 
                  smoothness_worst + compactness_worst + concavity_worst + 
                  concave.points_worst + symmetry_worst + fractal_dimension_worst, data=training_set_k, hidden = c(5,3), linear.output = FALSE)

p.nn.val <- compute(nn, testing_set_k[,1:30])

p <- sapply(p.nn.val$net.result,round)

val <-table(p, testing_set_k$diagnosis)
val.t <- sum(diag(val))/sum(val)
print(val.t)
```
